msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-11 paper.tex
Minor tweaks.

In human-like parameter regimes, \msprime's implementation of the
exact coalescent with recombination comprehensively outperforms
all other simulators, including those based on SMC
approximations~\citep{kelleher2016efficient}. However, it is important
to note that although the implementation of Hudson's algorithm is
very efficient, it is still quadratic in the scaled recombination rate
$\rho. This is because Hudson's algorithms tracks recombinations not only 
in segments ancestral to the sample, but also between ancestral segments.     
Estimates of the number of such "trapped" recombinations grow quadratically 
with genome length (\cite{Hein, Schierup, Wiuf}, Eq. 5.10). 
Because they do not track such recombinations, efficient SMC implementations 
will be faster that \msprime for sufficiently high recombination rate, 
genome length, or $N_e$. %Until we decide to approximate the recombination process
%between distant ancestral segment. Hint hint.

@@

@@

\subsection*{Discussion}
% Simulation is useful, and it's really easy for a single locus.
% This has lead to a culture of developing lots of simulators
% rather than focusing on one
Simulation has played a key role in the development of coalescent theory
and its use in understanding observed patterns of genetic variation.
The basic model can be simulated for $n$ samples in $O(n)$ time
and is straightforward to implement---for example, Hudson provided a fully
functional and highly efficient simulator as a C source code listing in an
appendix to his seminal review paper~\citep{hudson1990gene}. Because of
this ease, the dominant approach to producing simulators for extensions
to this basic model has been to develop them independently. This is
often seen as a learning opportunity, since those developing extensions
to the coalescent must have an excellent understanding of the basic
model in order to develop extensions to the coalescent.
The coalescent with recombination, however, is \emph{not} straightforward
to implement, and numerous different approaches have been
proposed over the
years~\citep{hudson1983properties,griffiths1997ancestral,wiuf1999recombination,
mcvean2005approximating}. It has only recently become clear that it is
possible to implement the model efficiently, and only through the use
of sophisticated data structures and an intricate and subtle
algorithm~\citep{kelleher2016efficient}.
This efficiency, and the ability to handle recombination in a principled
way is vital with today's vast numbers of whole genome samples.

@@

% Making lots of simulators has given us a fragmented and poort
% quality ecosystem.
The approach of developing ``in-house'' simulators
has lead to a huge proliferation of different programs, with
slightly different feature sets and often incompatible interfaces.
It is a fragmented ecosystem with significant issues with software
quality (see, e.g.,~\cite{yang2014critical});
few simulators are actively maintained after publication.
There seems to be little point
in continuing the huge duplication of effort we have seen.
It is not possible to achieve the performance levels of
\msprime\ without investing substantially in software engineering.
As demonstrated here, \msprime\ is extensible and we would argue
that developing this simulator as an open-source community asset
is of more benefit to all.

@@

% We need to acknowledge that for long genomes and for large population
% sizes msprime's performance does drop off and SMC simulators like
% scrm become more efficient.
Msprime's performance for large sample sizes is excellent, but
simulating species with large population sizes is still challenging.
An important avenue for future development is to improve performance
in these scenarios by incorporating data structures such as skip lists
and implementing more efficient coalescent approximations, as done
in \texttt{cosi2}~\citep{shlyakhter2014cosi2}.

In terms of mutation simulation, future work will need to do things like
Dawg~\citep{cartwright2005dna} and to implement
codon models and indels~\citep{fletcher2009indelible}.

>>>>

In human-like parameter regimes, \msprime's implementation of the
exact coalescent with recombination comprehensively outperforms
all other simulators, including those based on SMC
approximations~\citep{kelleher2016efficient}. However, it is important
to note that although the implementation of Hudson's algorithm is
very efficient, it is still quadratic in the scaled recombination rate
$\rho$. This is because Hudson's algorithms tracks recombinations not only
in segments ancestral to the sample, but also between ancestral segments.
Estimates of the number of such ``trapped'' recombinations grow quadratically
with genome length \citep[Eq.~5.10]{hein2004gene}.
Because they do not track such recombinations, efficient SMC implementations
will be faster that \msprime for sufficiently high recombination rate,
genome length, or $N_e$. %Until we decide to approximate the recombination process
%between distant ancestral segment. Hint hint.

@@

Things we can do: future work will need to do things like
Dawg~\citep{cartwright2005dna} and to implement models and
indels~\citep{fletcher2009indelible}.

@@

\subsection*{Discussion}
% Simulation is useful, and it's really easy for a single locus.
% This has lead to a culture of developing lots of simulators
% rather than focusing on one
Simulation has played a key role in the development of coalescent theory
and its use in understanding observed patterns of genetic variation.
The basic model can be simulated for $n$ samples in $O(n)$ time
and is straightforward to implement---for example, Hudson provided a fully
functional and highly efficient simulator as a C source code listing in an
appendix to the seminal review paper~\citep{hudson1990gene}. Because of
this ease, the dominant approach to producing simulators for extensions
to this basic model has been to develop them independently. This is
often seen as a learning opportunity, since those developing extensions
to the coalescent must have an excellent understanding of the basic
model in order to develop extensions to the coalescent.
The coalescent with recombination, however, is \emph{not} straightforward
to implement, and numerous different approaches have been
proposed over the
years~\citep{hudson1983properties,griffiths1997ancestral,wiuf1999recombination,
mcvean2005approximating}. It has only recently become clear that it is
possible to implement the model efficiently, and only through the use
of sophisticated data structures and
intricate algorithms~\citep{kelleher2016efficient}.

% Making lots of simulators has given us a fragmented and poor
% quality ecosystem.
The approach of developing ``in-house'' simulators
has lead to a huge proliferation of different programs, with
slightly different feature sets and often incompatible interfaces.
It is a fragmented ecosystem with significant issues with software
quality (see, e.g.,~\cite{yang2014critical}),
and few simulators are actively maintained after publication.
We have seen that \msprime\ (for larger samples at least)
has at worst comparable, but usually far superior
performance to more specialised methods.

% JK: I want to make the point that the established idea that
% developing your own simulation software is a good learning experience
% is pretty questionable these days, and tends to produce a lot of
% crappy unmaintained software. Not sure if the following paragraph
% does the job that well, though.

There seems to be little point in continuing the huge duplication of effort we
have seen. It is not possible to achieve the performance levels of \msprime\
without investing substantially in software engineering. As demonstrated here,
\msprime\ is extensible and we would argue that developing this simulator as an
open-source community asset is of more benefit to all. Implementing a
coalescent simulator may once have been a good way for (e.g.) a student to
learn about programming and simulation, but it is no longer reasonable to
expect this. There is simply too much engineering required. For example, as of
the 1.0.0 release \msprime\ consisted of around 13K lines of C and 11K
lines of Python, with test suites of 122 C test cases (7K lines of code) and
1350 Python test cases (22K lines of code), along with a suite of 217
statistical validation tests (6K lines of code). It requires a sustained effort
over years to produce well engineered, high-quality software. It may reasonably
be asked if students would learn more from contributing to an existing project
with established high-quality engineering standards, or from producing their own
software in a self-taught manner.

@@

@@
